Evolution of ORM Systems: Formal Foundations

### 2012-12-30 doc/description.tex
updated application operations typesetting

The model uses a set $Labels$ which contains all possible identifiers for the model. Each model component which has a label ($label \in Labels$) is unambiguously identified in the model.
\begin{gather*}
  Application = < Class* >
  \\ 
  app : Class* \rightarrow  Application \\ \\
  classes : Application \rightarrow Class* \\
  classes(app(cs)) = cs  \\\\
  class : Label \times Application \rightarrow Class   \\ 
  class(l, a) = c, c \in classes(a) \wedge name(c) = l \\ \\
  l \in Label, a = Application, cs \in Class* 
  \\ \\
  associated :  Label \times Application \rightarrow Boolean \\
  owningClass : Property \times Application \rightarrow Class  \\ owningClass(p, a) = c, c \in classes(a) \wedge p \in properties(c) \\
  owningClass : Association \times Application \rightarrow Class  \\
owningClass(as, a) = c, c \in classes(a) \wedge as \in associations(c)
\end{gather*}
The structure of an application consists not only of classes, it is defined by five other concepts. Besides the structure definition, there is defined a set of functions, which allows to manipulate the application model easily.

@@

\section{Database Model}
\label{sec:dbModel}
The relational database consists of database schema which defines the structure of the database and data, which in our ORM system represents instances. The database is:
$$
Database = <table*, row*>
$$
$table : Label \times Database \rightarrow Table $ \\
$table(l, d) = t, t \in tables(d) \wedge name(t) = l$ \\

\subsection{Database Schema Model}
The database schema model is defined by following concepts:
\paragraph{Table} represents a basic concept of database schema. It has a unique name, one or more column and it can be related to other tables in the schema by foreign keys. Rows in the table represents stored data.
$$
Table = (label, primaryKey, Column*, ForeignKey*)
$$
$name : Table \rightarrow Label $ \\
$name(< lab, pk, cols, fks)) = lab$ \\ \\
$primaryKey : Table \rightarrow PrimaryKey $ \\
$primaryKey(< lab, pk, cols, fks)) = pk$ \\ \\
$columns : Table \rightarrow Column* $ \\
$columns(< lab, pk, cols, fks)) = cols$ \\ \\
$foreignKeys : Table \rightarrow ForeignKey* $ \\
$foreignKeys(< lab, pk, cols, fks)) = fks$

\paragraph{Column} columns define possible data values and types which can be part of a table record.
$$
Column = (label, DbType, DefaultValue, Constraint*)
$$
$name : Column \rightarrow Label $ \\
$name(< lab, t, val, cons)) = lab $ \\ \\
$type : Column \rightarrow DbType $ \\
$type(< lab, t, val, cons)) = t $ \\ \\
$constraints : Column \rightarrow Constraint* $ \\
$constraints(< lab, t, val, cons)) = cons $ \\ \\
$owningTable : Column \times Database \rightarrow Table $ \\
$constraints(c, d) = tab, tab \in tables(d) \wedge c \in columns(d) $

\paragraph{Foreign key}
$$
ForeignKey = (label, tableRef, Constraint*)
$$
$name : ForeignKey \rightarrow Label $ \\
$name(< lab, tRef, cons)) = lab $ \\ \\
$reference : ForeignKey \rightarrow Label $ \\
$reference(< lab, tRef, cons)) = tRef $ \\ \\
$reference : ForeignKey \rightarrow Constraint* $ \\
$reference(< lab, tRef, cons)) = cons $ \\ \\
$owningTable : ForeignKey \times Database \rightarrow Table $ \\
$owningTable(fk, d) = tab, tab \in tables(d) \wedge fk \in foreignKeys(d) $

\paragraph{Primary key}
$$
PrimaryKey =  < label > 	
$$

@@

\paragraph{Constraint} there are two types of constraints defined in the model. Both constraints are column constraints - first constraint defines unique records in a column, second constraint defines non-empty columns.

\begin{center}
$Constraint = NOTNULL$ $|$ $UNIQUE $
\end{center}

\subsection{Model of Stored Data}

A row in a table is tuple consisting of reference to a concrete table and a set of value pairs, which represents concrete values of concrete columns.
$$row = < refTable, pair* >$$
$$pair = < reference, value >$$
The $reference $ is a name of a column, a foreign key or its primary key and $value$ represents concrete value stored in the database. Each row in a database has to refer to a table in database schema, otherwise the database is inconsistent. The consistency of database is then:
\begin{equation*}
\begin{gathered}
\forall r \in data(d), \exists t \in tables(d) : reference(r) = name(t) \\ \wedge \forall p \in pairs(r) \exists u \in columns(t) \cup foreignKeys(t) : reference(p) = name(u) 
\end{gathered}
\end{equation*}
There are two functions connected with stored data - first of them verify if the table contains any data:
$$contains : Table \times Database \rightarrow Boolean $$
\begin{equation*}
\begin{gathered}
	contains(t, d) = \begin{cases}
 \inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
  \inference{\forall r \in data(d) : refTable(r) \neq name(t)}{false}
 \end{cases} \\ \\
 t \in tables(d), d \in Database
\end{gathered}
\end{equation*}

@@

  \inference{cardinality(p) = 1 \wedge mandatory(p)
 \\ \wedge \exists t : t = table(name(owningClass(p)), d)}{\begin{gathered}
 	 columns(t) \cup column(name(p), \rho_t(type(p)), NOTNULL ) \\

\end{gathered}
} 
\\ \\

 \inference{ cardinality(p) > 1 \wedge ! mandatory(p) \\ \wedge \exists t : t = table(name(owningClass(p)), d)}{\begin{gathered}tables(d) \cup  table(name(p), primaryKey("Id"), \\ column("value", \rho_t(type(p)), \varnothing), \\ foreignKey(name(p), name(owningClass(p)), NOTNULL) \\

\end{gathered}}
\\ \\

@@

\subsubsection{Application Creation}
$$addClass: Class \times Application \rightarrow Application $$

$
addClass(c, a) = \inference{\forall s \in classes(A) : name(s) \neq name(c)}
{classes(a) \cup c}
$

$$AddProperty : Label \times Property \times Application \rightarrow Application $$

$
AddProperty(l, p, a) = \inference{\exists c \in classes(a) : name(c) = l \\ \wedge \forall q \in  properties(p) : name(q) \neq name(p)}{properties(c) \cup p }
$
$$AddAssociation : Label \times Association \times Application \rightarrow Application $$
$
AddAssociation(l, s, a) = \inference{\forall f \in associations(a) : name(f) \neq name(s) \\ \wedge \exists c \in classes(a) : name(c) = l \\ \wedge \exists k \in classes(a) : name(k) = reference(s) }{associations(c) \cup s}
$

\subsubsection{Application Deconstruction}
$$RemoveClass: Label \times Application \rightarrow Application $$
$
RemoveClass(q, a) = \inference{\exists c \in classes(a) : name(c) = q \\
\wedge not$ $instantiated(c) = \varnothing \wedge not$ $associated(c)}{classes(a) \setminus c }
$

$$RemoveProperty: Label \times Label \times Application \rightarrow Application $$

$
RemoveProperty(l_c, l_p, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists p \in properties(c) : name(p) = l_p}{properties(c) \setminus p }
$

$$RemoveAssociation : Label \times Labels \times Application \rightarrow Application $$

$
RemoveAssociation(l_c, l_a, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists s \in associations(c) : name(s) = l_a }{associations(c) \setminus s}
$

\subsubsection{Application Alternation}
\paragraph{Rename Class}
$$\rho_{RenC} : Label \times Label \times Application \rightarrow Application $$
\begin{equation*}
	\rho_{RenC}(l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \forall e \in classes(a) : name(e) \neq l_n}{\begin{gathered}
c.name = l_n \wedge \forall f \in associations(e), \\ e \in classes(a), reference(f) = l_c : f.reference = l_n 
\end{gathered}
}
\end{equation*}

\paragraph{Rename Property}
$$\rho_{RenP} : Label \times Label \times Label \times Application \rightarrow Application $$
\begin{equation*}
	\rho_{RenP}(l_c, l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists p \in properties(c) : name(p) = l_o \\ \wedge \forall e \in properties(c) : name(e) \neq l_n}{\begin{gathered}
p.name = l_n 
\end{gathered}
}
\end{equation*}

\paragraph{Rename Association}
$$\rho_{RenA} : Label \times Label \times Label \times Application \rightarrow Application $$
\begin{equation*}
	\rho_{RenP}(l_c, l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists a \in associations(c) : name(a) = l_o \\ \wedge \forall e \in associations(c) : name(e) \neq l_n}{\begin{gathered}
a.name = l_n 
\end{gathered}
}
\end{equation*}

>>>>

The model uses a set $Labels$ which contains all possible identifiers for the model. Each model component which has a label ($label \in Labels$) is unambiguously identified in the model.
\begin{gather*}
  Application = < Class* >
  \\ 
  app : Class* \rightarrow  Application \\ \\
  classes : Application \rightarrow Class* \\
  classes(app(cs)) = cs  \\\\
  class : Label \times Application \rightarrow Class   \\ 
  class(l, a) = c, c \in classes(a) \wedge name(c) = l \\ \\
  l \in Label, a = Application, cs \in Class* 
  \\ \\
  associated :  Label \times Application \rightarrow Boolean \\
  owningClass : Property \times Application \rightarrow Class  \\ owningClass(p, a) = c, c \in classes(a) \wedge p \in properties(c) \\
  owningClass : Association \times Application \rightarrow Class  \\
owningClass(as, a) = c, c \in classes(a) \wedge as \in associations(c) \\ \\
owningTable : Column \times Database \rightarrow Table  \\
owningTable(c, d) = tab, tab \in tables(d) \wedge c \in columns(d) \\ \\
owningTable : ForeignKey \times Database \rightarrow Table  \\
owningTable(fk, d) = tab, tab \in tables(d) \wedge fk \in foreignKeys(d)\\ \\ 
\end{gather*}
The structure of an application consists not only of classes, it is defined by five other concepts. Besides the structure definition, there is defined a set of functions, which allows to manipulate the application model easily.

@@

\section{Database Model}
\label{sec:dbModel}
The relational database consists of database schema which defines the structure of the database and data, which in our ORM system represents instances. The database is:
\begin{gather*}
Database = <table*, row*> \\ \\
db : Tables \times Rows \rightarrow Database \\ \\
tables : Database \rightarrow Tables \\
tables(db(tabs, rows)) = tabs \\ \\
data : Database \rightarrow Rows \\
data(db(tabs, rows)) = rows \\ \\
table : Label \times Database \rightarrow Table  \\
table(l, d) = t, t \in tables(d) \wedge name(t) = l \\ \\
tabs \in Tables, rows \in Rows, l \in Labels, d \in Databases
\end{gather*}

\subsection{Database Schema Model}
The database schema model is defined by following concepts:
\paragraph{Table} represents a basic concept of database schema. It has a unique name, one or more column and it can be related to other tables in the schema by foreign keys. Rows in the table represents stored data.
\begin{gather*}
Table = <label, primaryKey, Column*, ForeignKey*>\\ \\
tab : Label \times PrimaryKeys \times Columns \times ForeignKeys \rightarrow Table \\ \\
name : Table \rightarrow Label  \\
name(tab(l, pk, cols, fks)) = l \\ \\
primaryKey : Table \rightarrow PrimaryKeys  \\
primaryKey(tab(l, pk, cols, fks)) = pk \\ \\
columns : Table \rightarrow Columns  \\
columns(tab(l, pk, cols, fks)) = cols \\ \\
foreignKeys : Table \rightarrow ForeignKeys  \\
foreignKeys(tab(l, pk, cols, fks)) = fks \\ \\
l \in Label, pk \in PrimaryKey, cols \in Columns, fks \in ForeignKeys
\end{gather*}

\paragraph{Column} columns define possible data values and types which can be part of a table record.
\begin{gather*}
Column = <label, DbType, DefaultValue, Constraint*> \\ \\
col : Label \times DbType \times String \times Constraints \rightarrow Column \\ \\
name : Column \rightarrow Label  \\
name(col(l, t, val, cons)) = l  \\ \\
type : Column \rightarrow DbType  \\
type(col(l, t, val, cons)) = t  \\ \\
constraints : Column \rightarrow Constraints  \\
constraints(col(l, t, val, cons)) = cons  \\ \\
l \in Labels, t \in DbType, val \in Strings, cons \in Constraints
\end{gather*}

\paragraph{Foreign key}
\begin{gather*}
ForeignKey = <label, tableRef, Constraint*> \\ \\
fk : Label \times Label \times Constraints \rightarrow ForeignKey \\ \\
name : ForeignKey \rightarrow Label \\
name(fk(l, tRef, cons)) = lab  \\ \\
reference : ForeignKey \rightarrow Label  \\
reference(fk(l, tRef, cons)) = tRef  \\ \\
reference : ForeignKey \rightarrow Constraint*  \\
reference(fk(l, tRef, cons)) = cons  \\ \\
l, tRef \in Label, cons \in Constraints
\end{gather*}

\paragraph{Primary key}
\begin{gather*}
PrimaryKey =  < label > 	\\
pk : Label \rightarrow PrimaryKey
\end{gather*}

@@

\paragraph{Constraint} there are two types of constraints defined in the model. Both constraints are column constraints - first constraint defines unique records in a column, second constraint defines non-empty columns.
\begin{gather*}
Constraint = NOTNULL \; | \; UNIQUE 
\end{gather*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Model of Stored Data}
A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of reference to a concrete table and a set of value pairs, which represents concrete values of concrete columns.
\begin{gather*}
row = < refTable, pair* > \\
row : Label \times Pairs \rightarrow Row \\ \\
refTable : Row \rightarrow Label \\
refTable(row(l, pairs) = l \\ \\
contains : Table \times Database \rightarrow Boolean \\
contains(t, d) = \begin{cases}
 \inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
  \inference{\forall r \in data(d) : refTable(r) \neq name(t)}{false}
 \end{cases} \\ \\
 l \in Labels, pairs \in Pairs, t \in tables(d), d \in Database
\end{gather*}
There functions $contains$ verify if the table contains any data.

\begin{gather*}
pair = < refCol, value > \\
[Label \times Strings] \rightarrow Pair
\end{gather*}
The $reference $ is a name of a column, a foreign key or its primary key and $value$ represents concrete value stored in the database. Each row in a database has to refer to a table in database schema, otherwise the database is inconsistent. The consistency of database is then:
\begin{equation*}
\begin{gathered}
\forall r \in data(d), \exists t \in tables(d) : reference(r) = name(t) \\ \wedge \forall p \in pairs(r) \exists u \in columns(t) \cup foreignKeys(t) : reference(p) = name(u) 
\end{gathered}
\end{equation*}

@@

  \inference{cardinality(p) = 1 \wedge mandatory(p)
 \\ \wedge \exists t : t = table(name(owningClass(p)), d)}{\begin{gathered}
 	 columns(t) \cup column(name(p), \rho_t(type(p)), NOTNULL ) \\ 
\end{gathered}} 
\\ \\

 \inference{ cardinality(p) > 1 \wedge ! mandatory(p) \\ \wedge \exists t : t = table(name(owningClass(p)), d)}{\begin{gathered}tables(d) \cup  table(name(p), primaryKey("Id"), \\ column("value", \rho_t(type(p)), \varnothing), \\ foreignKey(name(p), name(owningClass(p)), NOTNULL) \\
\end{gathered}}
\\ \\

@@

\subsubsection{Application Creation}
\begin{gather*}
addClass: Class \times Application \rightarrow Application \\ 
addClass(c, a) = \inference{\forall s \in classes(a) : name(s) \neq name(c)}{classes(a) \cup c} \\ \\
c \in Classes, a \in Applications
\end{gather*}
\begin{gather*}
AddProperty : Label \times Property \times Application \rightarrow Application \\
AddProperty(l, p, a) = \inference{\exists c \in classes(a) : name(c) = l \\ \wedge \forall q \in  properties(p) : name(q) \neq name(p)}{properties(c) \cup p } \\
l \in Label, p \in Properties, a \in Applications
\end{gather*}
\begin{gather*}
AddAssociation : Label \times Association \times Application \rightarrow Application \\
AddAssociation(l, s, a) = \inference{\forall f \in associations(a) : name(f) \neq name(s) \\ \wedge \exists c \in classes(a) : name(c) = l \\ \wedge \exists k \in classes(a) : name(k) = reference(s) }{associations(c) \cup s}\\
l \in Labels, s \in Associations, a \in Applications
\end{gather*}

\subsubsection{Application Deconstruction}
\begin{gather*}
RemoveClass: Label \times Application \rightarrow Application  \\
RemoveClass(q, a) = \inference{\exists c \in classes(a) : name(c) = q \\
\wedge !\: instantiated(c) = \varnothing \wedge ! \: associated(c)} {classes(a) \setminus c } \\
q \in Label, a \in Applications
\end{gather*}
\begin{gather*}
RemoveProperty: Label \times Label \times Application \rightarrow Application \\
RemoveProperty(l_c, l_p, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists p \in properties(c) : name(p) = l_p}{properties(c) \setminus p }\\
l_c, l_p \in Label, a \in Applications 
\end{gather*}
\begin{gather*}
RemoveAssociation : Label \times Labels \times Application \rightarrow Application \\ 
RemoveAssociation(l_c, l_a, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists s \in associations(c) : name(s) = l_a }{associations(c) \setminus s} \\
l_c, l_a \in Labels, a \in Applications
\end{gather*}

\subsubsection{Application Alternation}
\paragraph{Rename Class}
\begin{gather*}
\rho_{RenC} : Label \times Label \times Application \rightarrow Application \\ 
\rho_{RenC}(l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \forall e \in classes(a) : name(e) \neq l_n}{\begin{gathered}
c.name = l_n \wedge \forall f \in associations(e), \\ e \in classes(a), reference(f) = l_c : f.reference = l_n 
\end{gathered}}\\
l_o, l_n \in Label, a \in Applications
\end{gather*}

\paragraph{Rename Property}
\begin{gather*}
\rho_{RenP} : Label \times Label \times Label \times Application \rightarrow Application \\ \\
	\rho_{RenP}(l_c, l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists p \in properties(c) : name(p) = l_o \\ \wedge \forall e \in properties(c) : name(e) \neq l_n}{\begin{gathered}
p.name = l_n 
\end{gathered}}\\
l_c, l_o, l_n \in Label, a \in Applications
\end{gather*}

\paragraph{Rename Association}
\begin{gather*}
\rho_{RenA} : Label \times Label \times Label \times Application \rightarrow Application \\
	\rho_{RenP}(l_c, l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists a \in associations(c) : name(a) = l_o \\ \wedge \forall e \in associations(c) : name(e) \neq l_n}{\begin{gathered}
a.name = l_n 
\end{gathered}}\\
l_c, l_o, l_n \in Label, a \in Applications
\end{gather*}
